Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality

### 2021-11-22 Code/Manuscript-SMOTE.Rmd
including multiple ways for determining presence

---
title: "Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality"
author: "Sarah J. Hart et al."
email: "sarah.hart@colostate.edu"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output: 
  bookdown::word_document2:
    reference_docx: WordTemplate.docx
    fig_caption: yes
    toc: no
    number_sections: no
    df_print: kable
csl: "`r here:::here('ecology.csl')`"
bibliography: "`r here:::here('MyLib.bib')`"
urlcolor: blue
linkcolor: blue
citationcolor: blue
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = F,
	message = FALSE,
	warning = FALSE,
	dpi = 300,
	progress = FALSE,
	cache = FALSE
)

@@

Key findings

-

@@

Paragraph 1: Theoretical context - plant & insect interactions

Many studies have sought to understand how biodiversity influences interactions between natural enemies and their resources. In the case of plant-herbivore interactions, greater plant diversity often reduces herbivory on plants at both the individual and population level [i.e., 'associational resistance', @barbosa_associational_2009]. This is hypothesized to occur when insect herbivores, particularly specialists, are less likely to find and stay in areas where their hosts are less abundant [i.e., 'resource concentration hypothesis', @root_organization_1973]. This may occur if insects are less likely to encounter their host due to decrease in host frequency or abundance [i.e. 'host dilution effect', @], heterospecific neighbors chemically mask host trees [i.e. 'host apparency hypothesis', @castagneryrol_effects_2014], and/or mixed stands favor greater abundance of herbivore natural enemies [i.e., 'enemies hypothesis', @elton_ecology_1985, @staab_influence_2020].

#However, increased plant diversity may also result in an increase in herbivore richness ['resource specialization hypothesis', @keddy_plant_1984].

Yet little research has examined community-level outcomes, which are often hard to quantify because of the number of potential interactions increases dramatically with increasing community diversity, particularly when natural enemies are generalists. Further, the effects of different natural enemies on their focal host populations often differ greatly, due to differences in susceptibility and mortality rates and the availability and quality (as viewed by their natural enemies) of resource communities. Critically, community-level outcomes may drive ecosystem processes, particularly when community diversity is low and plant-herbivore relationships are highly dynamic. In this context there is a particular need to understand the effects of tree diversity on outbreaks of irruptive insects, which in recent years (2003-2012) have affected 85 million hectares of forest globally, or ca. 18 million hectares more than wildfire [@van_lierop_global_2015], resulting in widespread changes in carbon sequestration [e.g., @kurz_mountain_2008], timber production [e.g. @schowalter_insect_2012], and water quality [e.g., @Mikkelson_bark_2013], among other things.

Paragraph 2: Bark beetles

Bark beetles (Curculionidae: Scolytinae) are among the few native insect species that can kill large numbers of trees in a single year. For instance in the western United States, recent outbreaks of native bark beetles have killed 3.8 billion trees [1997-2018; @hicke_characterizing_2020]. Bark beetles bore through the bark, where they mate and oviposit their eggs. Concurrently, bark beetles introduce pathogenic fungi Larvae feeding upon the phloem and fungal spread stop the translocation of water and nutrients and cause tree death. Conifer defense against bark beetles consists primarily of resin exudation that physically expels the beetle and allelochemicals, which repel or kill beetles. To overcome these defenses and colonize live trees, bark beetles rely on a mass-attack strategy, where pioneering beetles emit aggregation pheromones that call conspecifics to the tree. Typically bark beetles exist at low population levels and attack weakened trees, but as populations increase bark beetles attack increasingly better defended trees. Such severe mortality occurs only during outbreaks when pheromone-mediated mass-attack allows bark beetles to overcome tree defenses.

For instance, a primary goal of the Western Bark Beetle is to promote resilience of forests to bark beetle outbreaks by increasing the diversity of age classes and tree species.

Paragraph 3: Bark beetles

To better understand how community diversity influences interactions between bark beeetles and their resources, we use a natural system with inherently low tree and bark beetle diversity. Specifically our research focuses on tree mortality due to three bark beetle species, the mountain pine beetle (*Dendroctonus ponderosae*), spruce beetle (*D. rufipennis*), and western balsam bark beetle (*Dryocoetes confusus*), which in subalpine forests of Western North America predominantly attack lodgepole pine (*Pinus contorta*), Engelmann spruce (*Picea engelmannii*), and subalpine fir (*Abies lasiocarpa*), respectively. We use this system to ask:

(1) Is co-occurrence of multiple trees susceptible to bark beetle attack common?
(2) How often do stand conditions susceptible to multiple bark beetles occur?
(3) Given stand stand conditions susceptible to multiple bark beetles, does the  the occurrence or severity of cumulative bark beetle mortality

We expect that co-occurrence of trees of different host species will be common, but co-occurrence of stand conditions suitable for multiple bark beetle species will not. Given stands where conditions are suitable for multiple bark beetle species, occurrence of bark beetle-driven tree mortality will be greater (i.e. diversity begets diversity), but severity will be lower (i.e. resource concentration).

host tree richness or identity influence the occurrence and severity of tree mortality due to bark beetles at the community-scale?

To test these hypotheses, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/). Specifically we ask:

We also expected the severity of bark beetle infestation to vary with the number of agents present, but with two alternative hypotheses. If the population dynamics of each bark beetle species are independent, then stands with multiple agents will experience higher tree mortality than stands with only one agent (i.e. additive effects). Alternatively, lower tree mortality may occur if concurrent outbreaks of bark beetles of different species cause semiochemical confusion or if competitive release increases tree defensive capacity.

@@

The study area consists of subalpine lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) forest in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, Wyoming)

## Data
### FIA data

The FIA program is a single inventory program that includes all public and private forested land (>= 0.4 ha in size and >= 10% canopy cover) in the US. In the Western US, all plots are visited once every ten years [@gray_forest_2012]. The spatially and temporally distributed probabilistic sampling design is useful studies of the distribution of tree species [e.g., @iverson_predicting_1998; @rehfeldt_empirical_2006], forest insects [e.g., @derose_effects_2013], and tree mortality [e.g., @shaw_forest_2005]

At each FIA plot, field crews collect data for trees (>= 12.7 cm DBH) within four 7.32 m radius subplots arranged in a fixed pattern. Data on the proximate cause of death is collected for any tree (>= 12.7 cm DBH) that was alive at the previous visit and at revisit is dead using visible evidence (e.g., fire scars, bark beetle galleries).

@@

#### Determination of presence/absence of bark beetle activity

For each live tree (>= 12.7 cm DBH), field crews record up to three damaging agents, which are defined as agents that are likely to prevent the tree from surviving >2 years, reduce the growth of the tree in the near term, or negatively affect the tree's marketable products [@burrill_forest_2017].

Cause of death codes are very broad (e.g., "insect" or "disease"). Accuracy of FIA data is commonly assessed using blind checks, where two crews perform independent inventories. Agreement between mortality agent codes recorded in the two inventories is generally >80% [@anderegg_tree_2015]. Active damage is easier to identify, thus the codes for damaging agents are much more specific.

@@

#### Calculation of stand characteristics
For each plot, we then calculated
1 - basal area by host species
2 - quadratic mean diameter (QMD)
3 - basal area dominance (% total basal area) by species
4 - presence and severity of bark beetle activity (% total basal area) by bark beetle species

@@

@@

#### Selection of plots
We selected all plots that were part of the annual inventory and where all subplots were inventoried.

Within this plots, we characterized stand structure and composition using only live and recently killed (i.e. killed within the past ~10 years) trees greater that 12.7 cm DBH within the subplot.

```{r select plots}
df <- read.csv(here("Data/Processed/FIAdat-df.csv"))

@@

### Model the suitable stand conditions

However, benefits provided by mixtures are less evident for larger-scale disturbances [@jactel_tree_2021]
suggesting that changes in the structure of host communities, rather than biodiversity per se, can explain when a dilution effect should be observed.

Construct random forest models using synthetic minority oversampling technique (SMOTE), an approach for dealing with imbalanced data. The SMOTE approach oversamples the minority class by synthesizing new cases from the minority class. We compared this approach with two other common approaches for dealing with imbalanced data in a random forest framework - balanced and weighted random forests and selected the approach that classified affected plots with the greatest accuracy.

@@

```{r fit MPB RF model Balanced, eval=F}
response <- "agentMPB"
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
datBB.c <- df[df$focalhost=="Any",]

>>>>

---
title: "Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality"
author: "Sarah Hart"
email: "sarah.hart@colostate.edu"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output: 
  bookdown::word_document2:
    reference_docx: WordTemplate.docx
    fig_caption: yes
    toc: no
    number_sections: no
    df_print: kable
csl: "`r here:::here('ecology.csl')`"
bibliography: "`r here:::here('MyLib.bib')`"
urlcolor: blue
linkcolor: blue
citationcolor: blue
editor_options: 
  chunk_output_type: inline
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = F,
	message = FALSE,
	warning = FALSE,
	dpi = 300,
	progress = FALSE,
	cache = FALSE
)

@@

## Key findings

@@

Many studies have sought to understand how biodiversity influences interactions between natural enemies and their resources. In the case of plant-herbivore interactions, greater plant diversity often reduces the effects of herbivory [i.e., 'associational resistance', @barbosa_associational_2009], particularly when herbivores are specialists capable of rapid population growth [@otway_resource_2005]. While increases in plant diversity may decrease the effects of herbivore species on individuals and populations of plants, it is also hypothesized to increase herbivore richness ['resource specialization hypothesis', @keddy_plant_1984], which may result in more efficient resource use and greater depletion of the plant community, especially when herbivores are specialists [@cardinale_effects_2005; @finke_niche_2008]. Thus plant community-level outcomes of herbivory may depend upon both direct and indirect effects of plant diversity. Yet little research has examined the effects of plant diversity on community-level outcomes of herbivory, which critically may drive key ecosystem processes. This is particularly true in natural systems, where the number of potential interactions is high, the effects of different herbivores on their focal host populations often differ greatly, and the changes in plant diversity are often convoluted with changes in host quality [@jactel_tree_2021].

In the western United States, native bark beetles (Curculionidae: Scolytinae) have killed 3.8 billion trees over the past two decades [1997-2018; @hicke_characterizing_2020]. While there are more than 25 native bark beetle species, the mountain pine beetle (MPB; *Dendroctonus ponderosae*), spruce beetle (SB;  *Dendroctonus rufipennis*), and western balsam bark beetle (WBBB; *Dryocoetes confusus*) collectively have caused over 70% of the bark beetle-attributed tree mortality, most of which has occurred in high elevation forests [@hicke_chacterizing_2020].

Associational resistance is commonly hypothesized to occur due to several non-mutually exclusive mechanisms. First, herbivory may be lower in more diverse systems if herbivores are less likely to encounter, stay, and reproduce in areas where their hosts are less abundant [i.e., 'resource concentration hypothesis', @root_organization_1973]. Second, reductions in herbivory may occur if insects are less likely to find their hosts due to chemical or visual masking by heterospecific neighbors [i.e. 'host apparency hypothesis', @feeny_plant_1976, @sciebe_semiochemical_2011, @castagneryrol_effects_2014], Third, heterospecific neighbors may increase the abundance of herbivore natural enemies [i.e., 'enemies hypothesis', @elton_ecology_1985, @staab_influence_2020]. However, evidence for associational resistance in natural (i.e. not plantations) forested ecosystems.

Resource specialization 
Compensatory mechanisms
- only meaningful if populations susceptible to bb infestionat co- occur with tree populatoin ssucess to other

To test these hypotheses, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/).

We use this system to ask:

(1) How often do stand conditions suitable for multiple bark beetle species occur?
(2) Does the occurrence or severity of cumulative bark beetle activity increase with the host species richness?
(3) Does the co-occurrence of multiple bark beetle species result in more severe cumulative bark beetle mortality?

We expect that co-occurrence of trees of different host species will be common (i.e. resource specialization hypothesis), but co-occurrence of stand conditions suitable for multiple bark beetle species will not (i.e. associational resistance).

When stand conditions are suitable for multiple bark beetle species, we expect that at the population-scale occurrence of bark beetle-driven tree mortality wi

will be greater (i.e. resource specialization hypothesis), but severity will be lower (i.e. associational resistance). Alternatively, if

We also expected the severity of bark beetle infestation to vary with the number of agents present, but with two alternative hypotheses. If the population dynamics of each bark beetle species are independent, then stands with multiple agents will experience higher tree mortality than stands with only one agent. Alternatively, lower tree mortality may occur if concurrent outbreaks of bark beetles of different species cause semiochemical confusion or if competitive release increases tree defensive capacity.

@@

The study area consists of subalpine forests with lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) that are located in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, and Wyoming;  Fig. \@ref(fig:FigMap)). Our study area was constrained to the Intermountain West due to regional differences in data collection that exist within the FIA program [@burrill_forest_2017].

## Data

The FIA program inventories all public and private forested land using a spatially and temporally distributed sampling design, making it useful for the study of tree species distributions [e.g., @iverson_predicting_1998; @rehfeldt_empirical_2006] and tree demography [e.g., @shaw_forest_2005, @derose_effects_2013]. In the Western US, all plots are visited once every ten years [@gray_forest_2012]. At each plot, field crews collect data within four 7.32 m radius subplots arranged in a fixed pattern. For each tree within the plot, field crews record the species, size (diameter at breast height; DBH), and status (live or dead) for trees (>= 12.7 cm DBH).For each live tree, field crews additionally record up to three damaging agents (e.g. MPB), defined as any agent likely to prevent the tree from surviving >2 years, reduce the growth of the tree in the near term, or negatively affect the tree's marketable products [@burrill_forest_2017]. For dead trees that were alive at the previous visit, the proximate cause of death is also collected for any tree that was alive at the previous visit and at revisit is dead using visible evidence (e.g., fire scars, bark beetle galleries). Because identifying the mortality agent is more challenging cause of death codes are broad (e.g., "insect" or "disease").

We acquired all FIA data for the Intermountain West, but restricted our analyses to the most recent inventory for plots that contained a focal host species and where all subplots were inventoried. We then characterized stand structure and composition using only live and recently killed (i.e. killed within the past ~10 years) trees. each plot we calculated the basal area, quadratic mean diameter (QMD), and basal area dominance (% total basal area) for each host species. Additionally, we characterized the presence (i.e., at least one tree infested) and severity of bark beetle activity (% total basal area) by bark beetle species. For live Engelmann spruce, we listed the presence of SB when the agent was listed as a general bark beetle or SB specifically. Similarly, we listed the presence of MPB in lodgepole when agent was listed as a general bark beetle or MPB specifically. For live subalpine fir, we listed the presence of WBBB when the damaging was recorded as, subalpine fir decline, a general bark beetle, or WBBB specifically. Finally, we assumed the mortality of lodgepole pine, Engelmann spruce, and subalpine fir that was attributed an insect, was caused by the MPB, SB, and WBBB, respectively.

@@

@@

@@

#df$agentSB <- ifelse(df$stand.SB/(df$TPH.PIEN/6.018046/2.47105)>0.1, "SB", "absent")
#df$agentMPB <- ifelse(df$stand.MPB/(df$TPH.PICO/6.018046/2.47105)>0.1, "MPB", "absent")
#df$agentWBBB <- ifelse(df$stand.WBBB/(df$TPH.ABLA/6.018046/2.47105)>0.1, "WBBB", "absent")

@@

```{r select plots, eval=F}
df <- read.csv(here("Data/Processed/FIAdat-df.csv"))

@@

```{r maps all plots and hosts, eval=F}
df.all <- df[df$focalhost == "Any", ]
df.all$hostsFPlong <- df.all$hostsFP
df.all$agentsFplong <-factor(df.all$agentsF, levels=levels(df.all$agentsF), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB & SB", "MPB & WBBB", "SB & WBBB", "MPB, SB, & WBBB"))

map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=c(unaffected="#A6A6A6", MPB="#710002")) +tm_compass(position=c(0.1, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")

```{r FigMap, fig.cap="a caption", out.width = "504.8pt"}
knitr::include_graphics(here("Results/Figures/Figure-Map-Allplots.jpg"))
```

To determine how frequently co-occurrence of multiple bark beetle species was likely to occur, we first calculated the proportion of plots where multiple host species were present.

Because the effects of diversity may be convoluted with stand development processes and thus host size and availability, we limited our analyses to only the plots where host trees were susceptible to outbreak. To do so we used a random forest modeling framework [@breiman_random_2001], a distribution-free approach useful for modeling nonlinear relationships and complex interactions characteristic of ecological systems [@cutler_random_2007]. We constructed separate models for the presence of MPB, SB, and WBBB and as predictor variables included host QMD, host basal area, and host dominance, which have been commonly linked to increased risk of bark beetle activity [[e.g. REF MPB, REF SB, REF WBBB].  Because most stands in our dataset were unaffected by bark beetles, we used a synthetic minority oversampling technique (SMOTE) to deal the imbalance in the response [@chawla_smote_2002]. Briefly, the SMOTE approach oversamples the minority class by synthesizing new cases from the minority class (here affected stands). Models were fit in R [@r_core_team_r:_2020] using the **Tidymodels** package [@kuhn_tidymodels_2020]. We then assumed that each host species was susceptible to infestation when the predicted probability of occurrence of the focal bark beetle species was greater than 0.5.

```{r describe frequency of hosts and agents, eval=T}
## Hosts
hosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
hostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
hostgroups$hosts <- factor(hostgroups$hosts, levels=hostgroups$hosts, labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))

@@

suitablehostgroups <- bbo.sub %>% group_by(suitablehostsP) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
```

```{r create co-occurrence figure}
gghostgroups <- ggplot(hostgroups, aes(x=hosts, y=freq, fill=hosts))+geom_bar(stat="identity")+scale_fill_manual(values=c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"))+theme(legend.position="none")+xlab("")+ylab("frequency")

To determine if the proportion of plots affected by any bark beetle species varied with the number or identity of hosts present, we first used a proportion test to determine if groups differed followed by a pairwise proportion test to test for differences between groups. To account for multiple comparisons, we adjusted p-values using methods outlined by [@benjamini_false_1995].

```{r occurrence by hosts 2}
dfpred <- df
dfpred <- dfpred[dfpred$focalagent=="Any",]

res <- dfpred %>% count(hosts, affected) %>% group_by(hosts) %>% mutate(prop = prop.table(n))
x <- dfpred %>% count(hosts)

resa <- res[res$affected=="affected",]
resa$alln <- x$n

x <- prop.test(x=resa$n, n=resa$alln)
xx <- pairwise.prop.test(x=resa$n, n=resa$alln, p.adjust.method="BH")
xxp <-rstatix::cor_gather(xx$p.value)
xxp$var1 <- factor(xxp$var1, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$var2 <- factor(xxp$var2, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$Comparison <- apply(xxp[,c("var1", "var2")], 1, paste, collapse=" - ")
xxpr <- rcompanion::cldList(cor~ Comparison, data = xxp, threshold = 0.05)

resp <- res[res$affected=="affected",]
resp <- merge(resp, xxpr, by.x="hosts", by.y="Group")
resp <- resp[order(resp$prop),]
resp$Letter2 <- c("a", "b", "b", "c", "c", "d", "e")
resp$hosts <- factor(resp$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

ggoi <- ggplot(resp, aes(x=hosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave( here("Results/Figures/FigureOccurrencebHosts2.jpg"), plot=ggoi,  width=fig.width1, height=fig.width1*0.75,units="in")

```{r occurrence by suitable hosts 2}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[!dfpred$suitablehosts=="none",]
dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))

res <- dfpred %>% count(suitablehosts, affected) %>% group_by(suitablehosts) %>% mutate(prop = prop.table(n))
x <- dfpred %>% count(suitablehosts)

resa <- res[res$affected=="affected",]
resa$alln <- x$n

x <- prop.test(x=resa$n, n=resa$alln)
xx <- pairwise.prop.test(x=resa$n, n=resa$alln, p.adjust.method="BH")
xxp <-rstatix::cor_gather(xx$p.value)
xxp$var1 <- factor(xxp$var1, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$var2 <- factor(xxp$var2, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$Comparison <- apply(xxp[,c("var1", "var2")], 1, paste, collapse=" - ")
xxpr <- rcompanion::cldList(cor~ Comparison, data = xxp, threshold = 0.05)

resp <- res[res$affected=="affected",]
resp <- merge(resp, xxpr, by.x="suitablehosts", by.y="Group")
resp <- resp[order(resp$prop),]
resp$Letter2 <- c("a", "b", "c", "d", "d", "d", "e")
resp$suitablehosts <- factor(resp$suitablehosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

ggos <- ggplot(resp, aes(x=suitablehosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave( here("Results/Figures/FigureOccurrencebySuitablehosts2.jpg"), plot=ggos,  width=fig.width1, height=fig.width1*0.75,units="in")
```

```{r occurence combo}
ggo <- ggoi+theme(axis.title.x = element_blank(), axis.text.x = element_blank()) + ggos + plot_layout(ncol=1)

ggsave( here("Results/Figures/FigureOccurrencebyHosts.jpg"), plot=ggo,  width=fig.width1, height=fig.width1*1.3,units="in")

To determine if the severity of cumulative bark beetle-driven tree mortality was greater in stands with multiple susceptible hosts, we subsetted our data to only the plots affected by at least one bark beetle species. We then used Kruskall-Wallis test to determine if cumulative mortality differed with the identify of bark beetle species present. Followed by a Dunn test to determine if difference existed between groups, again accounting for false discovery rate using the Benjamini and Hochberg (1995) method.

```{r severity by hosts}
dfpred <- df
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affected=="affected",]
dfpred$hosts <- factor(dfpred$hosts, levels=levels(dfpred$hosts),  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

x <- kruskal.test(propTBA~hosts, data=dfpred)
DT <- dunnTest(propTBA~hosts, data=dfpred,method="bh")
res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)

res2 <- dfpred %>% group_by(hosts) %>% summarize(median=median(propTBA), max=max(propTBA))

res3 <- merge(res, res2, by.x="Group", by.y="hosts")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("a", "a", "b", "bc", "c", "c", "c")

ggsi <- ggplot(dfpred, aes(x=hosts, y=propTBA*100, fill=hosts))+geom_boxplot(notch=F)+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ ylab("severity (% BA)") +xlab("hosts")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/FigureSevertitybyhosts.jpg"), plot=ggsi, width=fig.width1,height=fig.width1,units="in")
```

```{r severity by suitable hosts 2}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affected=="affected",]
dfpred <- dfpred[!dfpred$suitablehosts=="none",]
dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1],  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

x <- kruskal.test(propTBA~suitablehosts, data=dfpred)
DT <- dunnTest(propTBA~suitablehosts, data=dfpred,method="bh")
res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)

res2 <- dfpred %>% group_by(suitablehosts) %>% summarize(median=median(propTBA), max=max(propTBA))

res3 <- merge(res, res2, by.x="Group", by.y="suitablehosts")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("a", "b", "c", "d", "cd", "d", "d")

ggss <- ggplot(dfpred, aes(x=suitablehosts, y=propTBA*100, fill=suitablehosts))+geom_boxplot(notch=F)+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ ylab("severity (% BA)") +xlab("suitable hosts")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/FigureSevertitybySuitablehosts.jpg"), plot=ggss, width=fig.width1,height=fig.width1,units="in")
```

```{r severity combo}
ggs <- ggsi+theme(axis.title.x = element_blank(), axis.text.x = element_blank()) + ggss + plot_layout(ncol=1)

ggsave( here("Results/Figures/FigureSeveritybyHosts.jpg"), plot=ggs,  width=fig.width1, height=fig.width1*1.3,units="in")
```

```{r determine co-occurrence of agents by hosts}
df.all <- df[df$focalagent=="Any",]

obsexp.dat <- as.data.frame(df.all  %>% group_by( agents) %>% summarise(n = n())%>% mutate(freq = n / sum(n)))
obsexp.dat$n.exp <- NA
obsexp.dat$freq.exp <- NA

df.all.MPB <- df.all[df.all$hosts %in% c("PICO+PIEN+ABLA", "PICO+PIEN", "PICO+ABLA"),] %>% group_by( agentMPB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

df.all.SB <- df.all[df.all$hosts %in% c("PICO+PIEN+ABLA", "PICO+PIEN", "PIEN+ABLA"),] %>% group_by( agentSB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

df.all.WBBB <- df.all[df.all$hosts %in% c("PICO+PIEN+ABLA", "PICO+ABLA", "PIEN+ABLA"),] %>% group_by( agentWBBB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

obsexp.dat[obsexp.dat$agents=="MPB+SB", ]$n.exp <- table(df.all$hosts)["PICO+PIEN"] * df.all.MPB[2,3]*df.all.SB[2,3]

obsexp.dat[obsexp.dat$agents=="MPB+WBBB", ]$n.exp <- table(df.all$hosts)["PICO+ABLA"] * df.all.MPB[2,3]*df.all.WBBB[2,3]

obsexp.dat[obsexp.dat$agents=="SB+WBBB", ]$n.exp <- table(df.all$hosts)["PIEN+ABLA"] * df.all.SB[2,3]*df.all.WBBB[2,3]

obsexp.dat[obsexp.dat$agents=="MPB+SB+WBBB", ]$n.exp <- table(df.all$hosts)["PICO+PIEN+ABLA"] * df.all.MPB[2,3]*df.all.SB[2,3]*df.all.WBBB[2,3]

obsexp.dat <- obsexp.dat[obsexp.dat$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB"), ]
obsexp.dat <- as.data.frame(obsexp.dat)
obsexp.dat2 <- data.frame(agents= c(obsexp.dat$agents,obsexp.dat$agents), n=c(obsexp.dat[,"n"], do.call(c,obsexp.dat[,"n.exp"])), variable=c(rep("observed",4),rep( "expected",4)))
```

```{r determine co-occurrence of agents by suitable hosts}
bbo.dat.ms <- bbo.dat[bbo.dat$focalagent=="Any" & bbo.dat$suitablehosts %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB"),]

obsexp.dat <- as.data.frame(bbo.dat.ms  %>% group_by( agents) %>% summarise(n = n())%>% mutate(freq = n / sum(n)))
obsexp.dat$n.exp <- NA
obsexp.dat$freq.exp <- NA

bbo.dat.ms.MPB <- bbo.dat[bbo.dat$focalagent=="Any" & bbo.dat$suitablehosts %in% c("MPB+SB+WBBB", "MPB+WBBB", "MPB+SB"),] %>% group_by( agentMPB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

bbo.dat.ms.SB <- bbo.dat[bbo.dat$focalagent=="Any" & bbo.dat$suitablehosts %in% c("MPB+SB+WBBB", "SB+WBBB", "MPB+SB"),] %>% group_by( agentSB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

bbo.dat.ms.WBBB <- bbo.dat[bbo.dat$focalagent=="Any" & bbo.dat$suitablehosts %in% c("MPB+SB+WBBB", "MPB+WBBB", "SB+WBBB"),] %>% group_by( agentWBBB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

obsexp.dat[obsexp.dat$agents=="MPB+SB", ]$n.exp <- table(bbo.dat.ms$suitablehosts)["MPB+SB"] * bbo.dat.ms.MPB[2,3]*bbo.dat.ms.SB[2,3]

obsexp.dat[obsexp.dat$agents=="MPB+SB", ]$freq.exp <- obsexp.dat[obsexp.dat$agents=="MPB+SB", ]$n.exp[[1]][1] /sum(obsexp.dat$n)

obsexp.dat[obsexp.dat$agents=="MPB+WBBB", ]$n.exp <- table(bbo.dat.ms$suitablehosts)["MPB+WBBB"] * bbo.dat.ms.MPB[2,3]*bbo.dat.ms.WBBB[2,3]

obsexp.dat[obsexp.dat$agents=="MPB+WBBB", ]$freq.exp <- obsexp.dat[obsexp.dat$agents=="MPB+WBBB", ]$n.exp[[1]][1] /sum(obsexp.dat$n)

obsexp.dat[obsexp.dat$agents=="SB+WBBB", ]$n.exp <- table(bbo.dat.ms$suitablehosts)["SB+WBBB"] * bbo.dat.ms.SB[2,3]*bbo.dat.ms.WBBB[2,3]

obsexp.dat[obsexp.dat$agents=="SB+WBBB", ]$freq.exp <- obsexp.dat[obsexp.dat$agents=="SB+WBBB", ]$n.exp[[1]][1] /sum(obsexp.dat$n)

obsexp.dat[obsexp.dat$agents=="MPB+SB+WBBB", ]$n.exp <- table(bbo.dat.ms$suitablehosts)["MPB+SB+WBBB"]*bbo.dat.ms.SB[2,3]*bbo.dat.ms.WBBB[2,3]*bbo.dat.ms.MPB[2,3]

obsexp.dat[obsexp.dat$agents=="MPB+SB+WBBB", ]$freq.exp <- obsexp.dat[obsexp.dat$agents=="MPB+SB+WBBB", ]$n.exp[[1]][1] /sum(obsexp.dat$n)

obsexp.dat <- obsexp.dat[obsexp.dat$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB"), ]
